---
title: "Systolic Blood Pressure (SBP) Dashboard"
format: 
  dashboard:
    page-title: "Systolic Blood Pressure (SBP) Dashboard"
theme: [sandstone, theme/custom.scss]

execute:
  freeze: false

editor_options: 
  chunk_output_type: console
---

```{r}
#| label: load-packages
#| message: false
#| include: false

library(nhanesA)
library(tidyverse)
library(Epi)
```

```{r}
#| label: load-data
#| message: false
#| include: false

# bmx_raw <- nhanes('P_BMX') |> tibble()
# bpxo_raw <- nhanes('P_BPXO') |> tibble()
# demo_raw <- nhanes('P_DEMO') |> tibble()

# saveRDS(bmx_raw, "data/P_BMX.Rds")
# saveRDS(bpxo_raw, "data/P_BPXO.Rds")
# saveRDS(demo_raw, "data/P_DEMO.Rds")

#bmx_raw <- readRDS("dashboard-main/data/P_BMX.Rds")
#bpxo_raw <- readRDS("dashboard-main/data/P_BPXO.Rds")
#demo_raw <- readRDS("dashboard-main/data/P_DEMO.Rds")

bmx_raw <- readRDS("data/P_BMX.Rds")
bpxo_raw <- readRDS("data/P_BPXO.Rds")
demo_raw <- readRDS("data/P_DEMO.Rds")

```


<!--
____________________________________________________________________________________________
DATA DEVELOPMENT AND CLEANING
____________________________________________________________________________________________-->

```{r}
#| label: inspect-data
#| message: false
#| include: false

dim(demo_raw)
dim(bmx_raw)
dim(bpxo_raw)

names(demo_raw)
names(bmx_raw)
names(bpxo_raw)
```

```{r}
#| label: merge-data
#| message: false
#| include: false

analytic_raw <-
  demo_raw |>
  select(
    SEQN,
    RIDSTATR,
    RIDAGEYR,
    RIAGENDR,
    DMDEDUC2
  ) |>
  left_join(
    bmx_raw |>
      select(
        SEQN,
        BMXBMI
      ),
    by = "SEQN"
  ) |>
  left_join(
    bpxo_raw |>
      select(
        SEQN,
        BPXOSY1,
        BPXOSY2,
        BPXOSY3
      ),
    by = "SEQN"
  ) |>
  filter(
    RIDSTATR == "Both interviewed and MEC examined",
    RIDAGEYR >= 21,
    RIDAGEYR <= 79
  )
```

```{r}
#| label: clean-data
#| message: false
#| include: false


analytic_data <-
  analytic_raw |>
  mutate(
    avg_sbp = rowMeans(
      cbind(BPXOSY1, BPXOSY2, BPXOSY3),
      na.rm = TRUE
    ),
    avg_sbp = if_else(is.nan(avg_sbp), NA_real_, avg_sbp),

    sex = factor(
      RIAGENDR,
      levels = c("Female", "Male")
    ),

    education = factor(
      case_when(
        DMDEDUC2 %in% c(
          "Less than 9th grade",
          "9-11th grade (Includes 12th grade with no diploma)"
        ) ~ "Less than HS",
        
        DMDEDUC2 == "High school graduate/GED or equivalent" ~ "HS Grad",
        
        DMDEDUC2 == "Some college or AA degree" ~ "Some College",
        
        DMDEDUC2 == "College graduate or above" ~ "College Grad",
        
        TRUE ~ NA_character_
      ),
      levels = c(
        "Less than HS",
        "HS Grad",
        "Some College",
        "College Grad"
      )
    ),

    obesity = factor(
      case_when(
        BMXBMI >= 30 ~ "Obese",
        BMXBMI < 30 ~ "Not obese",
        TRUE ~ NA_character_
      ),
      levels = c("Not obese", "Obese")
    ),
    
    high_sbp = factor(
      case_when(
        avg_sbp >= 120 ~ "Elevated/High SBP",
        avg_sbp < 120 ~ "Normal SBP",
        TRUE ~ NA_character_
        ),
      levels = c("Normal SBP", "Elevated/High SBP")
      )
  )
```

```{r}
#| label: check-data
#| message: false
#| include: false


analytic_data |>
  summarise(
    n = n(),
    complete_sbp = sum(!is.na(avg_sbp)),
    complete_bmi = sum(!is.na(BMXBMI)),
    complete_all = sum(
      complete.cases(
        SEQN,
        RIDSTATR,
        RIDAGEYR,
        avg_sbp,
        BMXBMI,
        sex,
        education,
        obesity
      )
    ),
    min_age = min(RIDAGEYR, na.rm = TRUE),
    max_age = max(RIDAGEYR, na.rm = TRUE),
    n_sbp_values = n_distinct(avg_sbp, na.rm = TRUE),
    n_bmi_values = n_distinct(BMXBMI, na.rm = TRUE)
  )
```



<!--
____________________________________________________________________________________________
DISPLAY
____________________________________________________________________________________________-->

#  {.sidebar}

This dashboard summarizes systolic blood pressure data among adults aged 21–79 years in the 2017 – March 2020 pre-pandemic dataset from NHANES.

| | |
|---|---|
| **Source** | NHANES |
| **Cycle** | 2017–March 2020 |
| **Population** | Adults aged 21–79 years |
| **Total participants** | 7,968 |
| **Complete data** | 7,187 |

------------------------------------------------------------------------
::: {.callout-note collapse="true"}
## Disclaimer

Because NHANES sampling weights were not incorporated for this course project, these results should not be interpreted as nationally representative estimates for all U.S. adults.
:::


# Overview

## Row {height="25%"}

```{r}
#| component: valuebox
#| title: "Participants"

list(
  value = nrow(analytic_data)
)
```

```{r}
#| component: valuebox
#| title: "Mean SBP"

mean_sbp_overall <- mean(analytic_data$avg_sbp, na.rm = TRUE)

list(
  color = "primary",
  value = paste(round(mean_sbp_overall, 1), "mmHg")
)
```

```{r}
#| component: valuebox
#| title: "Mean BMI"

mean_bmi_overall <- mean(analytic_data$BMXBMI, na.rm = TRUE)

list(
  color = "primary",
  value = paste(round(mean_bmi_overall, 1), "kg/m²")
)
```

```{r}
#| component: valuebox
#| title: "Age Range"

list(
  color = "primary",
  value = paste(
    min(analytic_data$RIDAGEYR, na.rm = TRUE),
    "–",
    max(analytic_data$RIDAGEYR, na.rm = TRUE),
    "years"
  )
)
```

## Row {height="75%"}

### Column {width="50%"}

```{r}
#| title: "Distribution of Systolic Blood Pressure"

analytic_data |>
  filter(!is.na(avg_sbp)) |>
  ggplot(aes(x = avg_sbp)) +
  geom_histogram(
    bins = 35,
    color = "white"
  ) +
  labs(
    x = "Average systolic blood pressure (mmHg)",
    y = "Number of participants"
  ) +
  theme_minimal(base_size = 18)
```

### Column {width="50%"}

```{r}
#| title: "Distribution of BMI"

analytic_data |>
  filter(!is.na(BMXBMI)) |>
  ggplot(aes(x = BMXBMI)) +
  geom_histogram(
    bins = 35,
    color = "white"
  ) +
  labs(
    x = "BMI (kg/m²)",
    y = "Number of participants"
  ) +
  theme_minimal(base_size= 18)
```

<!--
____________________________________________________________________________________________
Analysis B
____________________________________________________________________________________________
-->


# Sex

```{r}
#| label: analysis-b
#| results: hide
#| message: false
#| include: false

analysis_B <-
  analytic_data |>
  filter(
    !is.na(avg_sbp),
    !is.na(sex)
  )

analysis_B_summary <-
  analysis_B |>
  group_by(sex) |>
  summarise(
    n = n(),
    mean_sbp = mean(avg_sbp),
    sd_sbp = sd(avg_sbp),
    median_sbp = median(avg_sbp),
    IQR_sbp = IQR(avg_sbp),
    .groups = "drop"
  )

analysis_B_test <- t.test(
  avg_sbp ~ sex,
  data = analysis_B,
  conf.level = 0.90
)
```

## Row {height="25%"}

```{r}
#| component: valuebox
#| title: "Female Mean SBP"

female_mean <- analysis_B_summary |>
  filter(sex == "Female") |>
  pull(mean_sbp)

list(
  color = "primary",
  value = paste(round(female_mean, 1), "mmHg")
)
```

```{r}
#| component: valuebox
#| title: "Male Mean SBP"

male_mean <- analysis_B_summary |>
  filter(sex == "Male") |>
  pull(mean_sbp)

list(
  color = "primary",
  value = paste(round(male_mean, 1), "mmHg")
)
```

```{r}
#| component: valuebox
#| title: "Mean Difference"

mean_difference <- female_mean - male_mean

list(
  color = "primary",
  value = paste(round(mean_difference, 2), "mmHg")
)
```

## Row {height="75%"}

### Column {width="55%"}

```{r}
#| title: "SBP Distribution by Sex"

analysis_B |>
  ggplot(aes(x = sex, y = avg_sbp, fill = sex)) +
  geom_boxplot(alpha = 0.8, show.legend = FALSE) +
  labs(
    x = NULL,
    y = "Average systolic blood pressure (mmHg)"
  ) +
  theme_minimal(base_size = 18)
```

### Column {width="45%"}

```{r}
#| title: "Sex vs. SBP Summary"

analysis_B_summary |>
  mutate(
    mean_sbp = round(mean_sbp, 1),
    sd_sbp = round(sd_sbp, 1),
    median_sbp = round(median_sbp, 1),
    IQR_sbp = round(IQR_sbp, 1)
  ) |>
  knitr::kable(
    col.names = c(
      "Sex",
      "n",
      "Mean",
      "SD",
      "Median",
      "IQR"
    )
  )
```


<!--
____________________________________________________________________________________________
Analysis C
____________________________________________________________________________________________
-->

# Education {orientation="columns"}

```{r}
#| label: analysis-c
#| include: false

analysis_C <-
  analytic_data |>
  filter(
    !is.na(avg_sbp),
    !is.na(education)
  )

# Education level
analysis_C_summary <-
  analysis_C |>
  group_by(education) |>
  summarise(
    n = n(),
    mean_sbp = mean(avg_sbp),
    sd_sbp = sd(avg_sbp),
    median_sbp = median(avg_sbp),
    IQR_sbp = IQR(avg_sbp),
    se_sbp = sd_sbp / sqrt(n),
    lower_90 = mean_sbp - qt(0.95, df = n - 1) * se_sbp,
    upper_90 = mean_sbp + qt(0.95, df = n - 1) * se_sbp,
    .groups = "drop"
  )

# One-way ANOVA
analysis_C_model <-
  aov(avg_sbp ~ education, data = analysis_C)

# ANOVA results
analysis_C_anova <-
  summary(analysis_C_model)

# Tukey with 90% confidence intervals
analysis_C_tukey <-
  TukeyHSD(
    analysis_C_model,
    "education",
    conf.level = 0.90
  )

# Displaying Tukey output
analysis_C_tukey_table <-
  as.data.frame(analysis_C_tukey$education) |>
  rownames_to_column("comparison") |>
  rename(
    difference = diff,
    lower_90 = lwr,
    upper_90 = upr,
    p_value = `p adj`
  )

# Overall ANOVA p-value
analysis_C_p <-
  summary(analysis_C_model)[[1]][["Pr(>F)"]][1]
```

## Column {width="20%"}

```{r}
#| component: valuebox
#| title: "Participants"

list(
  color = "primary",
  value = nrow(analysis_C)
)
```

```{r}
#| component: valuebox
#| title: "College Grad Mean SBP"

lowest_education <-
  analysis_C_summary |>
  slice_min(mean_sbp, n = 1)

list(
  color = "primary",
  value = paste(
    round(lowest_education$mean_sbp, 1),
    "mmHg"
  )
)
```

```{r}
#| component: valuebox
#| title: "Less than HS Mean SBP"

highest_education <-
  analysis_C_summary |>
  slice_max(mean_sbp, n = 1)

list(
  color = "primary",
  value = paste(
    round(highest_education$mean_sbp, 1),
    "mmHg"
  )
)
```

```{r}
#| component: valuebox
#| title: "ANOVA p-value"

anova_p_label <-
  ifelse(
    analysis_C_p < 0.001,
    "< 0.001",
    format(round(analysis_C_p, 3), nsmall = 3)
  )

list(
  color = "primary",
  value = anova_p_label
)
```

## Column {width="40%"}

```{r}
#| title: "SBP Distribution by Education"

analysis_C |>
  ggplot(
    aes(
      x = education,
      y = avg_sbp,
      fill = education
    )
  ) +
  geom_boxplot(
    alpha = 0.8,
    show.legend = FALSE
  ) +
  labs(
    x = NULL,
    y = "Systolic blood pressure (mmHg)"
  ) +
  theme_minimal(base_size = 18) +
  theme(
    axis.text.x = element_text(
      angle = 20,
      hjust = 1
    )
  )
```


```{r}
#| title: "Mean SBP and 90% Confidence Intervals"

analysis_C_summary |>
  ggplot(
    aes(
      x = education,
      y = mean_sbp
    )
  ) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(
      ymin = lower_90,
      ymax = upper_90
    ),
    width = 0.15
  ) +
  labs(
    x = NULL,
    y = "Mean systolic blood pressure (mmHg)"
  ) +
  theme_minimal(base_size = 18) +
  theme(
    axis.text.x = element_text(
      angle = 20,
      hjust = 1
    )
  )
```

## Column {width="40%"}

### Group Summary

```{r}
#| title: "Education vs. SBP Summary"

analysis_C_summary |>
  select(
    education,
    n,
    mean_sbp,
    sd_sbp,
    median_sbp,
    IQR_sbp
  ) |>
  mutate(
    mean_sbp = round(mean_sbp, 1),
    sd_sbp = round(sd_sbp, 1),
    median_sbp = round(median_sbp, 1),
    IQR_sbp = round(IQR_sbp, 1)
  ) |>
  knitr::kable(
    col.names = c(
      "Education",
      "n",
      "Mean SBP",
      "SD",
      "Median",
      "IQR"
    )
  )
```

### Tukey Comparisons

```{r}
#| title: "Tukey HSD Pairwise Comparisons"

analysis_C_tukey_table |>
  mutate(
    difference = round(difference, 2),
    lower_90 = round(lower_90, 2),
    upper_90 = round(upper_90, 2),
    p_value = case_when(
      p_value < 0.001 ~ "<0.001",
      TRUE ~ sprintf("%.3f", p_value)
    )
  ) |>
  knitr::kable(
    col.names = c(
      "Comparison",
      "Mean Difference",
      "90% CI Lower",
      "90% CI Upper",
      "Adjusted p-value"
    )
  )
```



<!--
____________________________________________________________________________________________
Analysis D
____________________________________________________________________________________________
-->

# Obesity

```{r}
#| label: analysis-d
#| include: false

analysis_D <-
  analytic_data |>
  filter(
    !is.na(obesity),
    !is.na(high_sbp)
  )

# 2 x 2 table
# Rows: exposure
# Columns: outcome
analysis_D_table <-
  table(
    factor(
      analysis_D$obesity,
      levels = c("Obese", "Not obese")
    ),
    factor(
      analysis_D$high_sbp,
      levels = c("Elevated/High SBP", "Normal SBP")
    )
  )

# Run 2 x 2 analysis
analysis_D_results <-
  Epi::twoby2(
    analysis_D_table,
    conf.level = 0.90
  )

# Counts and within-obesity proportions for plotting
analysis_D_summary <-
  analysis_D |>
  count(obesity, high_sbp) |>
  group_by(obesity) |>
  mutate(
    proportion = n / sum(n)
  ) |>
  ungroup()

# Prevalence of elevated/high SBP
prevalence_obese <-
  analysis_D |>
  filter(obesity == "Obese") |>
  summarise(
    prevalence = mean(high_sbp == "Elevated/High SBP")
  ) |>
  pull(prevalence)

prevalence_not_obese <-
  analysis_D |>
  filter(obesity == "Not obese") |>
  summarise(
    prevalence = mean(high_sbp == "Elevated/High SBP")
  ) |>
  pull(prevalence)
```

## Row {height="25%"}

```{r}
#| component: valuebox
#| title: "Elevated/High SBP: Obese"

list(
  color = "primary",
  value = paste0(
    round(prevalence_obese * 100, 1),
    "%"
  )
)
```

```{r}
#| component: valuebox
#| title: "Elevated/High SBP: Not Obese"

list(
  color = "primary",
  value = paste0(
    round(prevalence_not_obese * 100, 1),
    "%"
  )
)
```

```{r}
#| component: valuebox
#| title: "Participants"

list(
  color = "primary",
  value = nrow(analysis_D)
)
```

```{r}
#| component: valuebox
#| title: "SBP Cutoff"

list(
  color = "primary",
  value = "≥120 mmHg"
)
```

## Row

### Column {width="50%"}

```{r}
#| title: "Elevated/High SBP by Obesity Status"

analysis_D_summary |>
  ggplot(
    aes(
      x = obesity,
      y = proportion,
      fill = high_sbp
    )
  ) +
  geom_col() +
  scale_y_continuous(
    labels = scales::percent
  ) +
  labs(
    x = NULL,
    y = "Proportion",
    fill = "SBP status"
  ) +
  theme_minimal(base_size = 18)
```

### Column {width="50%"}

```{r}
#| title: "2 × 2 Table"

analysis_D |>
  count(obesity, high_sbp) |>
  tidyr::pivot_wider(
    names_from = high_sbp,
    values_from = n
  ) |>
  knitr::kable(
    col.names = c(
      "Obesity Status",
      "Normal SBP",
      "Elevated/High SBP"
    )
  )
```

```{r}
#| title: "Epi 2 × 2 Results"

analysis_D_results
```




<!--
____________________________________________________________________________________________
Analysis E
____________________________________________________________________________________________
-->

# Education x Categorized SBP {orientation="columns"}

```{r}
#| label: analysis-e
#| include: false

analysis_E <- 
  analytic_data |>
  mutate(
    sbp_category = factor(
      case_when(
        avg_sbp < 120 ~ "Normal",
        avg_sbp >= 120 & avg_sbp < 130 ~ "Elevated",
        avg_sbp >= 130 & avg_sbp < 140 ~ "Stage 1 range",
        avg_sbp >= 140 ~ "Stage 2 range or higher",
        TRUE ~ NA_character_
      ),
      levels = c(
        "Normal",
        "Elevated",
        "Stage 1 range",
        "Stage 2 range or higher"
      )
    )
  ) |>
  filter(
    !is.na(education),
    !is.na(sbp_category)
  )

# Cross-tabulation
analysis_E_table <-
  table(
    analysis_E$education,
    analysis_E$sbp_category
  )

# Chi-squared test
analysis_E_chisq <-
  chisq.test(analysis_E_table)

# Counts and within-education proportions
analysis_E_summary <-
  analysis_E |>
  count(education, sbp_category) |>
  group_by(education) |>
  mutate(
    proportion = n/sum(n)
  ) |>
  ungroup()

# p-value
analysis_E_p <-
  analysis_E_chisq$p.value

# Expected counts for verifications
analysis_E_expected <-
  analysis_E_chisq$expected

```

## Column {width="20%"}
```{r}
#| component: valuebox
#| title: "Participants"

list(
  color = "primary",
  value = nrow(analysis_E)
)
```

```{r}
#| component: valuebox
#| title: "Education x SBP Categories"

list(
  color = "primary",
  value = paste(nlevels(analysis_E$education), "x", nlevels(analysis_E$sbp_category))
)
```

```{r}
#| component: valuebox
#| title: "Chi-square p-value"

chi_p_label <-
  ifelse(
    analysis_E_p < 0.001,
    "< 0.001"
  )

list(
  color = "primary",
  value = chi_p_label
)

```

``` {r}
#| component: valuebox
#| title: "Minimum Expected Count"

list(
  color = "primary",
  value = round(min(analysis_E_expected), 1)
)
```

## Column {width="40%"}

```{r}
#| title: "SBP Category by Education"

analysis_E_summary |>
  ggplot(
    aes(
      x = education,
      y = proportion,
      fill = sbp_category
    )
  ) +
  geom_col() +
  scale_y_continuous(
    labels = scales::percent
  ) + 
  labs(
    x = NULL,
    y = "Proportion",
    fill = "SBP category"
  ) +
  theme_minimal(base_size = 18) +
  theme(
    axis.text.x = element_text(
      angle = 20,
      hjust = 1
    )
  )

```

```{r}
#| title: "Distribution of SBP Categories"

analysis_E_summary |>
  ggplot(
    aes(
      x = sbp_category,
      y = proportion,
      fill = education
    )
  ) +
  geom_col(
    position = "dodge"
  ) + 
  scale_y_continuous(
    labels = scales::percent
  ) + 
  labs(
    x = NULL,
    y = "Proportion within education group",
    fill = "Education"
  ) +
  theme_minimal(base_size = 18) +
  theme(
    axis.text.x = element_text(
      angle = 20,
      hjust = 1
    )
  )
```

## Column {width="40%"}

### Row {height="33%"}

**Description:**
The chi-square test of independence showed evidence of an association between the education level and SBP categories, as the X² value computed to 109.67, with  p<0.001. Individuals with a higher education level tended to have a greater proportion of normal SBP values and a lower proportion of Stage 2 range or higher values. For example, 50.9% of college graduates were in the normal SBP category compared with 40.3% of participants with less than a high school education, while Stage 2 hypertension or crisis occurred in 13.1% and 25.8% of those groups, respectively.


### Row {height="40%"}

```{r}
#| title: "Education x SBP Category"

analysis_E |>
  count(education, sbp_category) |>
  tidyr::pivot_wider(
    names_from = sbp_category,
    values_from = n
  ) |>
  select(
    education,
    Normal,
    Elevated,
    'Stage 1 range',
    'Stage 2 range or higher'
  ) |>
  knitr::kable(
    col.names = c(
      "Education",
      "Normal",
      "Elevated",
      "Stage 1 range",
      "Stage 2 range or higher"
    )
  )
```

### Chi-square Test

### Row {height="27%"}

```{r}
#| title: "Chi-square Test"

tibble(
  Statistic = unname(analysis_E_chisq$statistic),
  'Degrees of Freedom' = unname(analysis_E_chisq$parameter),
  `p-value` = analysis_E_chisq$p.value
) |>
  mutate(
    Statistic = round(Statistic, 2),
    `p-value` = case_when(
      `p-value` < 0.001 ~ "<0.001",
      TRUE ~ sprintf("%.3f", `p-value`)
    )
  ) |>
  knitr::kable()
```
